---
title: "Kamath CD83_BCAS2 Validation"
output: html_notebook
author: "Zena Chatila (zkc2001@cumc.columbia.edu)"
output:
  html_document: 
    toc: true
    number_sections: true
editor_options:
  chunk_output_type: console
---
Purpose: The purpose of this notebook is to:
1) Integrate the microglia from our dataset with the microglia from the Kamath et al (2022) dataset
2) Idenfiy microglia carrying the CD83_BCAS2 signature
3) Compute proportions of miroglia carrying the CD83_BCAS2 signature in Parkinson's disease versos control samples in the Kamath data

# Load packages
```{r}
set.seed(0)
library(DT)
library(knitr)
library(Matrix)
library(patchwork)
library(ggplot2)
library(cowplot)
library(ggpubr)
library(Seurat)
library(dplyr)
library(tidyverse)
library(harmony)
setwd("/path/to/kamath_integration")
```

# LOAD DATASETS
```{r}
## load Kamath et al dataset
Kamath_obj <- readRDS("kamath_integrated_microglia.rds")
DefaultAssay(Kamath_obj) <- "integrated"
Idents(Kamath_obj) <- "Cell_Type"
Kamath_obj2=subset(Kamath_obj, idents=c("Macro_CD200R1","MG_MKI67"),invert=T)

##load Chatila dataset
micro_zena <- readRDS("micro_defined_taxonomy.rds")
micro_zena$cluster_name <- micro_zena$final_names
micro_zena$cluster_name <- gsub("^", "Zena_", micro_zena$cluster_name)

# Create a new column by joining 'Region' and 'disease_status'
new_column <- paste(micro_zena$Region, micro_zena$Disease, sep = "_")

table(micro_zena$Region)
table(micro_zena$cluster_name)

Idents(micro_zena) <- "Region"
Zena_subset=subset(micro_zena, idents="SN")

Idents(Zena_subset) <- "cluster_name"
Zena_subset2=subset(Zena_subset, idents=c("Zena_PVM","Zena_Mono"),invert=T)
```

# INTEGRATE BOTH DATASETS
```{r}
Kamath_obj2$dataset <- "Kamath"
Zena_subset2$dataset <- "Zena"
obj <- merge(Kamath_obj2, Zena_subset2 )
DefaultAssay(obj) <- "RNA"
ifnb.list <- SplitObject(obj, split.by = "dataset")

# normalize and identify variable features for each dataset independently
ifnb.list <- lapply(X = list(Kamath_obj2,Zena_subset2), FUN = function(x) {
  DefaultAssay(x) <- "RNA"
  x <- NormalizeData(x)
  x <- FindVariableFeatures(x, selection.method = "vst", nfeatures = 2000)
})


features <- SelectIntegrationFeatures(object.list = ifnb.list)
anchors <- FindIntegrationAnchors(object.list = ifnb.list, anchor.features = features)
obj.combined <- IntegrateData(anchorset = anchors)
DefaultAssay(obj.combined) <- "integrated"


obj2=ScaleData(obj.combined)
obj2=FindVariableFeatures(obj2)
obj2=RunPCA(obj2)

obj2=RunHarmony(obj2,group.by="dataset",dims=1:20)
obj2=RunUMAP(obj2,reduction="harmony",dim=1:20)
obj2=FindNeighbors(obj2,reduction="harmony",dims=1:20)
obj2=FindClusters(obj2,resolution=.6)
obj2$cluster_name2 <- obj2$cluster_name
obj2$cluster_name2[obj2$dataset =="Kamath"] <- NA

obj2$Cell_Type2 <- obj2$Cell_Type
obj2$Cell_Type2[obj2$dataset =="Zena"] <- NA

obj2=FindClusters(obj2,resolution=.3)

saveRDS(obj2, file = "initial_merged_objects_v2.rds")
```

# PLOT DATA
```{r}
dp1 <- (DimPlot(obj2,label=T)+DimPlot(obj2, group.by="dataset", label=T))/
  (DimPlot(obj2, group.by="cluster_name", label=T)+DimPlot(obj2, group.by="Cell_Type", label=T))
ggsave(paste0("UMAP_Kamath_Zena_integrated_dims=20.png"), dp1, height=12, width=14)

key_marks <- read.csv("micro_final_markers_wilcox_names.csv")

dp <- DotPlot(obj2, features= key_marks %>%
                dplyr::filter(final_names=="CD83_BCAS2", p_val_adj <= .05, pct.1-pct.2>=.4) %>% arrange(desc(pct.1-pct.2)) %>% dplyr::pull(gene),
              cols = c("blue","lightgrey", "firebrick1"))+
  scale_color_gradient2(low="blue", mid="lightgrey",high="firebrick1")+
  coord_flip()

png("CD83_BCAS2_markers_0.3.png", height=5,width=8,units="in", res=900)
print(dp)
dev.off()

fp <- FeaturePlot(obj2, features= key_marks %>%
                    dplyr::filter(final_names=="CD83_BCAS2", p_val_adj <= .05, pct.1-pct.2>=.4) %>% arrange(desc(pct.1-pct.2)) %>% dplyr::pull(gene) %>%
                    head(9),
                  ncol=3,cols=c("lightblue","lightgrey","firebrick1"))
ggsave(paste0("Feature_plot_Kamath_Zena_integrated_no_preds_dims=20_0.3.png"),
       fp, height=12, width=14)


markers <- FindAllMarkers(obj2,pos.only=T)
markers %>% group_by(cluster) %>%filter(pct.1-pct.2 >= .4)%>%  dplyr::summarize(n = dplyr::n())
markers %>% write.csv("integrated_markers_0.4.csv",quote=F, row.names=F)

dp <- DotPlot(obj2, features= markers %>%
                dplyr::filter(cluster=="7", p_val_adj <= .05, pct.1-pct.2>=.3) %>% arrange(desc(pct.1-pct.2)) %>% dplyr::pull(gene),
              cols = c("blue","lightgrey", "firebrick1"))+
  scale_color_gradient2(low="blue", mid="lightgrey",high="firebrick1")+
  coord_flip()
png("/mnt/vast/hpc/homes/jm4454/PD_Lipids/ATAC/0203_CD83_BCAS2_markers_0.3_cluster_7_top_markers.png", height=5,width=8,units="in", res=900)
print(dp)
dev.off()


obj_kamath <- subset(obj2, subset = dataset=="Kamath")
obj_zena <- subset(obj2, subset = dataset=="Zena")

dp <- DotPlot(obj_kamath, features= key_marks %>%
                dplyr::filter(final_names=="CD83_BCAS2", p_val_adj <= .05, pct.1-pct.2>=.4) %>% arrange(desc(pct.1-pct.2)) %>% dplyr::pull(gene),
              cols = c("blue","lightgrey", "firebrick1"))+
  scale_color_gradient2(low="blue", mid="lightgrey",high="firebrick1")+
  coord_flip()
png("CD83_BCAS2_markers_0.3_kamath_data.png", height=5,width=8,units="in", res=900)
print(dp)
dev.off()

dp <- DotPlot(obj_zena, features= key_marks %>%
                dplyr::filter(final_names=="CD83_BCAS2", p_val_adj <= .05, pct.1-pct.2>=.4) %>% arrange(desc(pct.1-pct.2)) %>% dplyr::pull(gene),
              cols = c("blue","lightgrey", "firebrick1"))+
  scale_color_gradient2(low="blue", mid="lightgrey",high="firebrick1")+
  coord_flip()
png("CD83_BCAS2_markers_0.3_zena_data.png", height=5,width=8,units="in", res=900)
print(dp)
dev.off()
```

# Calculate Kamath CD83_BCAS2 Proportions
## Load meta data
```{r}
data_file <- "kamath.tsv.gz"
if (file.exists(data_file)) {
  meta <- vroom::vroom(data_file)
} else {
  kamath <- readRDS("initial_merged_objects_v2.rds")
  vroom::vroom_write(kamath@meta.data, data_file)
}
```

## Compute props
```{r}
props_kamath <- meta %>%
  group_by(donor_id) %>%
  filter(dataset == "Kamath") %>%
  filter(n() > 100) %>%
  ungroup() %>%
  mutate(across(matches("integrated_snn_res.0.|donor_id"), as.character)) %>%
  mutate(integrated_snn_res.0.3 = paste0("cluster_", integrated_snn_res.0.3)) %>%
  mutate(donor_id = paste0("donor_", donor_id)) %>%
  dplyr::select(all_of(c("integrated_snn_res.0.3", "Disease",
                         "Sex", "Donor_Age", "donor_id"))) %>%
  get_props("donor_id", "integrated_snn_res.0.3") %>%
  mutate(Disease = as.factor(Disease),
         Disease = factor(Disease, levels = c("NPC", "PD")),
         Sex = as.factor(Sex),
         Sex = factor(Sex, levels = c("M", "F")))
glimpse(props_kamath)

stopifnot(nrow(props_kamath) == 196 & ncol(props_kamath) == 7)

props_kamath %>%
  mutate(across(all_of(c("donor_id", "Disease", "Sex")), as.factor),
         across(starts_with("integrated.RNA_snn_res.0."), as.factor)) %>%
  datatable()

# Define cell types
cell_types <- unique(props_kamath$integrated_snn_res.0.3)
cell_types <- structure(cell_types, names = cell_types)

```

## Prop analysis for Kamath
• Cluster 7 carries the CD83_BCAS2 signature
```{r}
# SN comparison: robust GLM of SN
comp <- purrr::map_dfr(cell_types, function(cell_type) {
  output <- robust_glm(prop ~ Disease + Donor_Age + Sex,
                       subset(props_kamath, integrated_snn_res.0.3 == cell_type),
                       family = "quasibinomial")
  tidy_terms(output, exponentiate = F)
}, .id = "cell_type") %>%
  group_by(term) %>%
  mutate(fdr = p.adjust(p.value_hc, "fdr")) %>%
  ungroup()

# how many clusters are enriched/depeleted in disease?
comp %>%
  mutate(cell_type = as.factor(cell_type), term = as.factor(term)) %>%
  datatable()
```
