---
title: "Martirosyan Validation"
output: html_notebook
author: "Zena Chatila (zkc2001@cumc.columbia.edu)"
output:
  html_document: 
    toc: true
    number_sections: true
editor_options:
  chunk_output_type: console
---

Purpose: The purpose of this notebook is to:
1) Integrate the microlgial nuclei from our dataset with all microlial nuclei in the Martirosyan et al. (2024) dataset
2) Identify microglia in the Smajic et al. dataset carrying the CD83_BCAS2 signature
3) Perform proportion analysis to identify whether CD83_BCAS2 proportions differ in Parkinson's versus control donors from the Smajic et al. dataset.

# Set Up
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Packages
```{r}
library(DT)
library(knitr)
library(Matrix)
library(patchwork)
library(Seurat)
library(dplyr)
library(harmony)
library(ggplot2)
```

# Load Martirosyan data
```{r}
Martirosyan_toplevel <- readRDS("<path_to_datasetA>/input/toplevel_Martirosyan_object.rds")

head(Martirosyan_toplevel@meta.data)
table(Martirosyan_toplevel$broadclass)
Assays(Martirosyan_toplevel)

Idents(Martirosyan_toplevel) <- "broadclass"
DefaultAssay(Martirosyan_toplevel) <- "RNA"
Martirosyan_micro <- subset(Martirosyan_toplevel, idents = "Micro")
```

# Extract and save counts from Assay5 for the Microglia subset
```{r}
counts <- GetAssayData(Martirosyan_micro[["RNA"]], slot = "counts")
meta <- Martirosyan_micro@meta.data

# Save counts and metadata
saveRDS(counts, file="Martirosyan_counts.RDS")
saveRDS(meta, file="Martirosyan_meta.RDS")
```

# Load and process Martirosyan object
```{r}
Martirosyan_meta <- readRDS("Path/To/Directory/Martirosyan_meta.RDS")
Martirosyan_counts <- readRDS("Path/To/Directory/Martirosyan_counts.RDS")

Martirosyan_obj <- CreateSeuratObject(Martirosyan_counts, meta.data = Martirosyan_meta)
Martirosyan_obj$Region <- "Martirosyan_SNpc"
Martirosyan_obj$dataset <- "Martirosyan_study"
Martirosyan_obj$Region_disease <- paste("SNpc", Martirosyan_obj$Clinical.diagnosis, sep = "_")
Martirosyan_obj$donor_id <- Martirosyan_obj$Sample.ID
Martirosyan_obj$Sample.ID <- NULL
Martirosyan_obj$Brain.Bank.ID <- NULL

head(Martirosyan_obj@meta.data)

DefaultAssay(Martirosyan_obj) <- "RNA"
```

# Load and process chatila object
```{r}
chatila_obj <- readRDS("Path/To/Directory/ChatilaObject.rds")
head(chatila_obj@meta.data)

# Create a new column by joining 'Region' and 'Disease'
new_column <- paste(chatila_obj$Region, chatila_obj$Disease, sep = "_")
chatila_obj <- AddMetaData(object = chatila_obj, metadata = new_column, col.name = "Region_disease")
chatila_obj$chatila_cluster_name <- chatila_obj$final_names

chatila_obj@assays$RNA@layers$data <- NULL
chatila_obj@assays$RNA@layers$scale.data <- NULL

chatila_obj$donor_id <- chatila_obj$Donor
chatila_obj$Donor <- NULL
chatila_obj$dataset <- "Columbia"

head(chatila_obj@meta.data)

print("chatila_obj prepared")
```

# Combine the Martirosyan and chatila objects
```{r}
combined_obj <- merge(Martirosyan_obj, chatila_obj)
combined_obj <- NormalizeData(combined_obj)
combined_obj <- FindVariableFeatures(combined_obj)
combined_obj <- ScaleData(combined_obj)
combined_obj <- RunPCA(combined_obj)
print("Normalisation done")
```

# Integrate data using Harmony
```{r}
combined_obj <- IntegrateLayers(object = combined_obj, method = HarmonyIntegration, orig.reduction = "pca", new.reduction = "harmony", verbose = FALSE)

combined_obj <- FindNeighbors(object = combined_obj, dims = 1:20, reduction = "harmony")
combined_obj <- FindClusters(object = combined_obj, resolution = c(0.2, 0.3, 0.4, 0.5))
combined_obj <- RunUMAP(object = combined_obj, reduction = "harmony", dims = 1:20)

# Save combined object
saveRDS(combined_obj, file="umap_bydataset_Martirosyan_chatila_lognorm_harmony_micro.rds")
```

# Plot and save the results
```{r}
pdf("bydonor_20pcs_lognorm_microglia_AY_Martirosyan.pdf", useDingbats = F)
p1 = ElbowPlot(combined_obj, ndims = 30); plot(p1)
p1 = DimPlot(combined_obj, reduction = "umap", group.by = "donor_id") + NoLegend(); plot(p1)
p1 = DimPlot(combined_obj, reduction = "umap", group.by = "donor_id"); plot(p1)
p1 = DimPlot(combined_obj, reduction = "umap", group.by = "dataset"); plot(p1)
p1 = DimPlot(combined_obj, reduction = "umap", split.by = "dataset"); plot(p1)
p1 = DimPlot(combined_obj, reduction = "umap", group.by = "Region_disease"); plot(p1)
p1 = DimPlot(combined_obj, reduction = "umap", group.by = "chatila_cluster_name", label = T); plot(p1)
p1 = DimPlot(combined_obj, reduction = "umap", group.by = "RNA_snn_res.0.2", label = T); plot(p1)
dev.off()
T
```

# Load metadata
```{r}
# Load and explore data
data_file <- "mol_degen.tsv.gz"
if (file.exists(data_file)) {
  meta <- vroom::vroom(data_file)
} else {
  mol_degen <- readRDS("umap_bydataset_Martirosyan_chatila_lognorm_harmony_micro")
  vroom::vroom_write(mol_degen@meta.data, data_file)
}
```

# Compute props for Martirosyan data
```{r}
props_mol_degen <- meta %>%
  group_by(donor_id) %>%
  filter(dataset == "Martirosyan_study") %>%
  filter(n() > 100) %>%
  ungroup() %>%
  mutate(across(starts_with("RNA_snn_res.0."), as.character)) %>%
  mutate(RNA_snn_res.0.4 = paste0("cluster_", RNA_snn_res.0.3)) %>%
  mutate(donor_id = paste0("donor_", donor_id)) %>%
  dplyr::select(all_of(c("RNA_snn_res.0.4", "Clinical.diagnosis",
                         "Sex", "Age", "donor_id"))) %>%
  get_props("donor_id", "RNA_snn_res.0.4") %>%
  mutate(Clinical.diagnosis = as.factor(Clinical.diagnosis),
         Clinical.diagnosis = factor(Clinical.diagnosis,
                                     levels = c("Control", "Parkinson's")),
         Sex = as.factor(Sex),
         Sex = factor(Sex, levels = c("female", "male")))
glimpse(props_mol_degen)

stopifnot(nrow(props_mol_degen) == 364 & ncol(props_mol_degen) == 7)

props_mol_degen %>%
  mutate(across(starts_with("RNA_snn_res.0."), as.factor)) %>%
  datatable()

# Define cell types
cell_types <- unique(props_mol_degen$RNA_snn_res.0.4)
cell_types <- structure(cell_types, names = cell_types)
```

# Prop analysis for Martirosyan data
• Cluster 11 at res 0.4 carries the CD83_BCAS2 signature
```{r}
# SN comparison: robust GLM of SN
comp <- purrr::map_dfr(cell_types, function(cell_type) {
  output <- robust_glm(prop ~ Clinical.diagnosis + Age + Sex,
                       subset(props_mol_degen, RNA_snn_res.0.4 == cell_type),
                       family = "quasibinomial")
  tidy_terms(output, exponentiate = F)
}, .id = "cell_type") %>%
  group_by(term) %>%
  mutate(fdr = p.adjust(p.value_hc, "fdr")) %>%
  ungroup()

# how many clusters are enriched/depeleted in disease?
comp %>%
  mutate(cell_type = as.factor(cell_type), term = as.factor(term)) %>%
  datatable()
```
