---
title: "Merge micro batches"
author: "Zena Chatila (zkc2001@cumc.columbia.edu)"
output:
  html_document:
    df_print: paged
    toc: TRUE
    number_sections: TRUE
editor_options:
  chunk_output_type: inline
---
# Description
This notebook converges microglial nuclei from the old and new batches of samples from our PD dataset into one seurat V5 object. This combined seurat object can then be used for integration and clustering of the microglial nuclei in the dataset. 

# Set up
#### Load packages and set seed
```{r message = F, warning = F}
set.seed(123)
library(sparseMatrixStats)
library(Matrix)
library(ggplot2)
library(DT)
library(magrittr)
library(dplyr)
library(Seurat)
```

#### General options
```{r warning = F, message = F}
options(dplyr.summarise.inform = F, DT.options = list(
  pageLength = 10, language = list(search = 'Filter:'), extensions = 'Buttons',
  filter = 'top', buttons = 'csv', autoWidth = T, scrollX = T, scrollY = T
  ))
```

# Compile master seurat object accross old and new
## Load data
```{r}
# Load microglial objects for batch one and batch two
load("microglia_clustering_for_paper_20230316.RData")
load("new_microglia_clustering_20240311.RData")

#rename old and new microglial objects
micro_old <- microglia_merged_obj_v4
micro_new <- new_mic_obj_v2
rm(microglia_merged_obj_v4, new_mic_obj_v2)

stopifnot(nrow(micro_old) == nrow(micro_new))
# How many cells per object
sapply(c(micro_old, micro_new), ncol)

# Load covariates
clinical_data <- read.csv("PDFMetaData.csv")
donor_data <- read.csv("donor_metadata.csv")
neuron_counts <- read.csv("NeuronCounts.csv") #microscopic neuromelanized neuron neuron counts from the SN
```

## Merge counts and metadata into a new seurat object
> NOTE: Micro new counts are not concatenated across samples. We need to compile them into a gene by cells_across_samples matrix.

```{r}
# Counts
counts_old <- micro_old[["RNA"]]$counts
counts_new <- JoinLayers(micro_new)[["RNA"]]$counts
counts_raw <- cbind(counts_old, counts_new)

# Metadata
metadata_raw <- bind_rows(list("old" = micro_old[["sample_ID"]],
                               "new" = micro_new[["sample_ID"]]),
                          .id = "batch")

rm(micro_old, micro_new, counts_old, counts_new)

# Add covariates
metadata_raw %<>%
  left_join(clinical_data, by = "sample_ID") %>%
  left_join(donor_data, by = "Donor") %>%
  left_join(neuron_counts, by = "Donor")

# Format covariates
metadata_raw %<>%
  mutate(Donor = paste0("donor_", Donor),
         Region = as.factor(Region),
         Region = factor(Region, c("SN", "VTA", "SI", "HypoTs")),
         NL = as.character(NL),
         NL = as.factor(paste0("NL", NL)),
         NL = factor(NL, levels = c("NL0", "NL1", "NL2", "NL3", "NL4")),
         LB = as.factor(LB),
         LB = factor(LB, levels = c("NONE", "LOW", "MED", "HIGH")),
         BRAAKLB = as.factor(BRAAKLB),
         BRAAKLB = factor(BRAAKLB, levels = c("IV", "V", "VI", "0")),
         Sex = case_when(grepl("M", Sex) ~ "M", grepl("F", Sex) ~ "F",
                         TRUE ~ NA_character_),
         Sex = as.factor(Sex),
         Sex = relevel(Sex, "M"),
         batch = as.factor(batch),
         Disease = as.factor(Disease),
         Disease = relevel(Disease, "NPC")) %>%
  select(-all_of(c("TID", "Batch2")))

stopifnot(nrow(metadata_raw) == ncol(counts_raw) & nrow(metadata_raw) == 84477)

# Create seurat object
micro <- CreateSeuratObject(counts_raw, meta.data = metadata_raw)
rm(counts_raw, metadata_raw)
stopifnot(ncol(micro) == 84477 & nrow(micro) == 22177)
micro
```

# Donor info
```{r message = F, warning = F}
micro@meta.data %>%
  group_by(Donor, Region, batch) %>%
  summarize(num_cells = n(), median_nCount_RNA = median(nCount_RNA, na.rm = T),
            median_nFeature_RNA = median(nFeature_RNA, na.rm = T),
            across(all_of(c("Disease", "NL", "LB", "Sex", "Age")), first)) %>%
  datatable()
```

# Expression info
```{r}
counts <- micro[["RNA"]]$counts
mean_gene_expr <- rowMeans(counts)
var_gene_expr <- sparseMatrixStats::rowVars(counts)
gene_info <- cbind(mean_gene_expr, var_gene_expr)

# Gene distribution
ggplot(gene_info, aes(mean_gene_expr, after_stat(scaled))) +
  geom_density() +
  theme_classic()

# Expression vs variance relationship
ggplot(gene_info, aes(log10(1 + mean_gene_expr), sqrt(var_gene_expr))) +
  geom_point() +
  theme_classic()
rm(counts, mean_gene_expr, var_gene_expr, gene_info)
```

# Save
```{r}
# Save if it doesn't exist
if (!file.exists("micro_merged.rda")) saveRDS(micro, "micro_merged.rda")
rm(micro)
```

# Sessioninfo
```{r}
sessionInfo()
```
