---
title: "Proportion Analysis for Parkinson's disease and Control Regions"
author: "Zena Chatila (zkc2001@cumc.columbia.edu)"
output:
  html_document: 
    toc: true
    number_sections: true
editor_options:
  chunk_output_type: console
---
# Description
The purpose of this notebook is to calculate proportions of microglial subpopulations in each tissue sample, and to investigate whether there are disease associated differences in proportions, across regions.

# Set Up
## Load packages
```{r message = F}
options(DT.options = list(pageLength = 10, extensions = 'Buttons',
                          buttons = 'csv', autoWidth = T, scrollX = T,
                          scrollY = T))

set.seed(0)
library(DT)
library(knitr)
library(Matrix)
library(patchwork)
library(forcats)
library(ggplot2)
library(cowplot)
library(ggpubr)
#library(SeuratDisk)
library(Seurat)
library(purrr)
library(tidyr)
library(dplyr)
# source get_props function for calculating proportions
source("./get_props.R")
# source robust_glm function for proportion analysis
source("./robust_glm.R")
# Add column filters to all DT::my_datatable tables
my_datatable <- function(...) datatable(..., filter = list(position = 'top',
                                                           clear = F))
```

# Load and explore data
```{r message = F}
data_file <- "micro_define_taxonomy_meta_data.tsv.gz"
if (file.exists(data_file)) {
  meta <- vroom::vroom(data_file)
} else {
  micro <- readRDS("micro_defined_taxonomy.rda")
  vroom::vroom_write(micro@meta.data, data_file)
}

# basic stats
stopifnot(nrow(meta) == 68559)
stopifnot(ncol(meta) == 26)
glimpse(meta)

```

# Compute proportions
```{r, fig.dim = c(10, 10)}
# filtering and computing props
final_names_levels <- c(
  "ZBTB16_NHSL1", "TMEM163_HAMP", "NAALADL2", "RPS11_FTL", "NPAS3_DLG2",
  "P2RY12_CX3CR1", "BAG3_CHORDC1", "CD83_BCAS2", "PTPRG_GLDN", "MKI67",
  "PVM", "Mono"
)

props <- meta %>%
  group_by(sample_ID) %>%
  filter(n() > 100) %>%
  ungroup() %>%
  mutate(across(starts_with("RNA_snn_res.0."), as.character),
         final_clusters = as.character(final_clusters),
         Donor = as.character(Donor),
         batch = as.character(batch),
         NL = as.character(NL)) %>%
  select(-all_of(c("seurat_clusters", "orig.ident", "nCount_RNA",
                   "nFeature_RNA", "optimal_res"))) %>%
  get_props("sample_ID", "final_names") %>%
  mutate(Region = as.factor(Region),
         Region = factor(Region,
                         levels = c("SN", "VTA", "SI", "HypoTs")),
         Disease = as.factor(Disease),
         Disease = factor(Disease, levels = c("NPC", "PD")),
         Sex = as.factor(Sex),
         Sex = factor(Sex, levels = c("M", "F")),
         NL = as.factor(NL),
         NL = factor(NL, levels = c("0", "1", "2", "3", "4")),
         LB = as.factor(LB),
         LB = factor(LB, levels = c("NONE", "LOW", "MED", "HIGH")),
         BRAAKLB = as.factor(BRAAKLB),
         BRAAKLB = factor(BRAAKLB, levels = c("IV", "V", "VI", "0")))
glimpse(props)

stopifnot(nrow(props) == 1644 & ncol(props) == 23)

#Check if any clusters are only present in cases or controls
props %>%
  group_by(Region, Disease, final_names) %>% 
  summarize(n = sum(prop > 0)) %>%
  pivot_wider(id_cols = c("Region", "final_names"), names_from = Disease,
              values_from = n) %>%
  my_datatable()

props %>%
  mutate(across(all_of(c("sample_ID", "Donor", "batch", "Region", "Sex",
                         "Disease")), as.factor),
         across(starts_with("RNA_snn_res.0."), as.factor)) %>%
  my_datatable()

# Define cell types
cell_types <- unique(props$final_names)
cell_types <- structure(cell_types, names = cell_types)
```

# Visualize proportions
```{r}
# plot props of all cell types in PD and NPCs (all regions)
props %>%
   mutate(final_names = as.factor(final_names),
          final_names = factor(final_names, levels = final_names_levels)) %>%
  ggplot(aes(Region, sqrt(prop), color = Disease)) +
    geom_boxplot(outlier.shape = NA, coef = 0, width = .7,
                 position = position_dodge(width = .85)) +
    ggbeeswarm::geom_beeswarm(dodge.width = .85, corral = "gutter", alpha = 0.3,
                              corral.width = 0.3, method = "swarm", cex = 1,
                              priority = "density", size = .8) +
    scale_color_manual(values = c(NPC = "#45CAFF", PD = "#FF1B6B")) +
    facet_wrap(~ final_names) +
    labs(x = "Region", y = "sqrt( Relative Proportion )", color = "Diagnosis") +
    theme_classic() +
    theme(panel.grid = element_blank(),
          axis.line = element_blank(),
          legend.position = "top",
          axis.ticks = element_blank())

## props_distribution for PD and NPC
ggplot(props,
       aes(sqrt(prop), after_stat(scaled), group = as.factor(final_clusters),
           color = as.factor(final_clusters))) +
  geom_density() +
  facet_grid(Disease ~ Region) +
  labs(title = "Distribution of sqrt(Relative Proportion)",
       x = "Props", y = "Density", color = "Cluster") +
  theme_classic()
```


### Save proportion file
```{r}
write.csv(props, file = "pd_analysis_props.csv")
```

# Proportion differences across regions per diagnosis
## NPC
```{r}
props_npc <- props %>%
  filter(Disease == "NPC") %>%
  droplevels()

region_per_cell_type_npc <- purrr::map_dfr(cell_types, function(cell_type) {
  # filtering to each cluster
  props_npc_cell_type <- props_npc %>%
    filter(final_names == cell_type)
  
  # define  model for assessing region
  region_diff_npc <- nlme::lme(
    sqrt(prop) ~ Region + Sex + Age + batch,
    random = ~ 1 | Donor,
    data = props_npc_cell_type
  )
  
  # create output table with the effect of region
  broom.mixed::tidy(region_diff_npc, effects = "fixed", conf.int = T)
}, .id = "cell_type") %>%
  group_by(term) %>%
  mutate(fdr = p.adjust(p.value, method = "fdr"),
         cell_type = as.factor(cell_type), term = as.factor(term)) %>%
  select(cell_type, term, estimate, statistic, p.value, fdr, everything()) %>%
  ungroup()

my_datatable(region_per_cell_type_npc)
```

Findings at FDR < 0.05: No differences in microglial proportions across control regions

## PD
```{r}
props_pd <- props %>%
  filter(Disease == "PD") %>%
  droplevels()

region_per_cell_type_pd <- purrr::map_dfr(cell_types, function(cell_type) {
  # filtering to each cluster
  props_pd_cell_type <- props_pd %>%
    filter(final_names == cell_type)
  
  # define model for assessing region
  region_diff_pd <- nlme::lme(
    sqrt(prop) ~ Region + Sex + Age + batch,
    random = ~ 1 | Donor,
    data = props_pd_cell_type
  )
  
  # create output table with the effect of region
  broom.mixed::tidy(region_diff_pd, effects = "fixed", conf.int = T)
}, .id = "cell_type") %>%
  group_by(term) %>%
  mutate(fdr = p.adjust(p.value, method = "fdr"),
         cell_type = as.factor(cell_type), term = as.factor(term)) %>%
  select(cell_type, term, estimate, statistic, p.value, fdr,
         everything()) %>%
  ungroup()

my_datatable(region_per_cell_type_pd)
```

Findings at FDR < 0.05 across PD regions: 
• NPAS3_DLG2 is increased in SI compared to SN
• NPAS3_DLG2 is increased in HypoTs compared to SN
• RPS11_FLT and NAALADL2 are decreased in SI compared to SN

# Proportion differences between NPC and PD per region
## SN
```{r}
# filter to props SN
props_sn <- props %>%
  filter(Region == "SN") %>%
  droplevels()

# SN comparison: robust GLM of SN
sn_comp <- purrr::map_dfr(cell_types, function(cell_type) {
  output <- robust_glm(prop ~ Disease + Age + Sex + batch,
                       subset(props_sn, final_names == cell_type),
                       family = "quasibinomial")
  tidy_terms(output, exponentiate = T)
}, .id = "cell_type") %>%
  group_by(term) %>%
  mutate(fdr = p.adjust(p.value_hc, "fdr")) %>%
  ungroup()

# how many clusters are enriched/depeleted in disease?
sn_comp %>%
  mutate(cell_type = as.factor(cell_type), term = as.factor(term)) %>%
  my_datatable()
```

Findings at FDR < 0.05:
• CD83_BCAS2 is reduced in PD SN compared to NPC SN


### What if we stratify by sex?
#### Males only
```{r fig.width = 12, fig.fullwidth = TRUE}
# filter to props SN
props_sn_M <- props_sn %>%
  filter(Sex == "M") %>%
  droplevels()
glimpse(props_sn_M)

# SN comparison: robust GLM of SN
sn_M_comp <- purrr::map_dfr(cell_types, function(cell_type) {
  output <- robust_glm(prop ~ Disease + Age + batch,
                       subset(props_sn_M, final_names == cell_type),
                       family = "quasibinomial")
  tidy_terms(output, exponentiate = T)
}, .id = "cell_type") %>%
  group_by(term) %>%
  mutate(fdr = p.adjust(p.value_hc, "fdr")) %>%
  ungroup()

# how many clusters are enriched/depeleted in disease?
sn_M_comp %>%
  mutate(cell_type = as.factor(cell_type), term = as.factor(term)) %>%
  my_datatable()
```

Findings at FDR < 0.05:
In males: 
• Age is associated with less P2RY12_CX3CR1 homeostatic microglia
• Disease is associated with decreased CD83_BCAS2


#### Females only
```{r}
# filter to props SN
props_sn_F <- props_sn %>%
  filter(Sex == "F") %>%
  droplevels()
glimpse(props_sn_F)

# SN comparison: robust GLM of SN
sn_F_comp <- purrr::map_dfr(cell_types, function(cell_type) {
  output <- robust_glm(prop ~ Disease + Age + batch,
                       subset(props_sn_F, final_names == cell_type),
                       family = "quasibinomial")
  tidy_terms(output, exponentiate = T)
}, .id = "cell_type") %>%
  group_by(term) %>%
  mutate(fdr = p.adjust(p.value_hc, "fdr")) %>%
  ungroup()

# how many clusters are enriched/depeleted in disease?
sn_F_comp %>%
  mutate(cell_type = as.factor(cell_type), term = as.factor(term)) %>%
  my_datatable()
```


### Are there any age specific effects? Let's stratify by disease and look at age

#### PD only 
```{r}
# filter to props SN
props_sn_pd <- props_sn %>%
  filter(Disease == "PD") %>%
  droplevels()
glimpse(props_sn_pd)

# SN comparison: robust GLM of SN
sn_pd_comp <- purrr::map_dfr(cell_types, function(cell_type) {
  output <- robust_glm(prop ~ Age + Sex + batch,
                       subset(props_sn_pd, final_names == cell_type),
                       family = "quasibinomial")
  tidy_terms(output, exponentiate = T)
}, .id = "cell_type") %>%
  group_by(term) %>%
  mutate(fdr = p.adjust(p.value_hc, "fdr")) %>%
  ungroup()

# how many clusters are enriched/depeleted in disease?
sn_pd_comp %>%
  mutate(cell_type = as.factor(cell_type), term = as.factor(term)) %>%
  my_datatable()
```

Findings at FDR < 0.05:
• in PD, RPS11_FLT is associated with Female sex

#### NPCs only
```{r}
# filter to props SN
props_sn_npc <- props_sn %>%
  filter(Disease == "NPC") %>%
  droplevels()
glimpse(props_sn_npc)

# SN comparison: robust GLM of SN
sn_npc_comp <- purrr::map_dfr(cell_types, function(cell_type) {
  output <- robust_glm(prop ~ Age + Sex + batch,
                       subset(props_sn_npc, final_names == cell_type),
                       family = "quasibinomial")
  tidy_terms(output, exponentiate = T)
}, .id = "cell_type") %>%
  group_by(term) %>%
  mutate(fdr = p.adjust(p.value_hc, "fdr")) %>%
  ungroup()

# how many clusters are enriched/depeleted in disease?
sn_npc_comp %>%
  mutate(cell_type = as.factor(cell_type), term = as.factor(term)) %>%
  my_datatable()
```

Findings with FDR < 0.05
• Age is associated with reduced P2RY12_CX3CR1 in NPCs
• BAG3_CHORDC1 is associated with F in NPCs

## VTA
```{r}
# filter to props VTA
props_vta <- props %>%
  filter(Region == "VTA") %>%
  droplevels()
glimpse(props_vta)

# vta comparison: robust GLM of SN
cell_types <- unique(props_vta$final_names)
cell_types <- structure(cell_types, names = cell_types)
vta_comp <- purrr::map_dfr(cell_types, function(cell_type) {
  output <- robust_glm(prop ~ Disease + Sex + Age + batch,
                       subset(props_vta, final_names == cell_type),
                       family = "quasibinomial")
  tidy_terms(output, exponentiate = T)
}, .id = "cell_type") %>%
  group_by(term) %>%
  mutate(fdr = p.adjust(p.value_hc, "fdr")) %>%
  ungroup()

# how many clusters are enriched/depeleted in disease?
vta_comp %>%
  mutate(cell_type = as.factor(cell_type), term = as.factor(term)) %>%
  my_datatable()
```

Findings FDR < 0.05:
• BAG3_CHORDC1 is increased in PD compared to NPC VTA 
• NAALADL2, P2RY12, and RPS11_FLT are decreased in PD compared to NPC VTA


## SI
```{r}
props_si <- props %>%
  filter(Region == "SI") %>%
  droplevels()
glimpse(props_si)

# SI comparison: robust GLM of SI
cell_types <- unique(props_si$final_names)
cell_types <- structure(cell_types, names = cell_types)
si_comp <- purrr::map_dfr(cell_types, function(cell_type) {
  output <- robust_glm(prop ~ Disease + Sex + Age + batch,
                       subset(props_si, final_names == cell_type),
                       family = "quasibinomial")
tidy_terms(output, exponentiate = T)
}, .id = "cell_type") %>%
  group_by(term) %>%
  mutate(fdr = p.adjust(p.value_hc, "fdr")) %>%
  ungroup()

# how many clusters are enriched/depeleted in disease?
si_comp %>%
  mutate(cell_type = as.factor(cell_type), term = as.factor(term)) %>%
  my_datatable()
```

Findings at FDR < 0.05:
No clusters are associated with disease status in SI

## HypoTs
```{r}
props_hypots <- props %>%
  filter(Region == "HypoTs") %>%
  droplevels()
glimpse(props_hypots)

# HypoT comparison: robust GLM of HypoT
cell_types <- unique(props_hypots$final_names)
cell_types <- structure(cell_types, names = cell_types)
hypots_comp <- purrr::map_dfr(cell_types, function(cell_type) {
  output <- robust_glm(prop ~ Disease + Sex + Age + batch,
                       subset(props_hypots, final_names == cell_type),
                       family = "quasibinomial")
  tidy_terms(output, exponentiate = T)
}, .id = "cell_type") %>%
  group_by(term) %>%
  mutate(fdr = p.adjust(p.value_hc, "fdr")) %>%
  ungroup()

# how many clusters are enriched/depleted in disease?
hypots_comp %>%
  mutate(cell_type = as.factor(cell_type), term = as.factor(term)) %>%
  my_datatable()
```

Findings at FDR < 0.05:
No clusters are associated with disease status in HypoTs

# Sessioninfo
```{r}
sessionInfo()
```
