---
title: "Siletti Validation"
output: html_notebook
author: "Zena Chatila (zkc2001@cumc.columbia.edu)"
output:
  html_document: 
    toc: true
    number_sections: true
editor_options:
  chunk_output_type: console
---

Purpose: The purpose of this notebook is to:
1) Integrate the microlgial nuclei from controls in our dataset with all microlial nuclei in the Siletti et al. (2022) dataset
2) Identify microglia in the Siletti et al. dataset carrying the CD83_BCAS2 signature
3) Perform proportion analysis to identify regional differences in proportions of CD83_BCAS2 throughout the non-diseased human brain. To do so, we compared proprotions in each siletti region against the proportion of CD83_BCAS2 in the substantia nigra (SN) from controls within our dataset. 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages
```{r message = F}
options(DT.options = list(
  pageLength = 10, language = list(search = 'Filter:'), extensions = 'Buttons',
  filter = 'top', buttons = 'csv', autoWidth = T, scrollX = T, scrollY = T
  ))

set.seed(0)
library(DT)
library(knitr)
library(Matrix)
library(patchwork)
library(ggplot2)
library(cowplot)
library(ggpubr)
library(Seurat)
library(harmony)
library(dplyr)
select <- dplyr::select
# source robust_glm function for proportion analysis
source("./robust_glm.R")
```

# Load and preprocess Chatila microglia data
```{r}
# Load and preprocess chatila_etal microglia data
# ------------------------------
cat("Loading chatila_etal microglia data...\n")
chatila_micro <- readRDS("micro_defined_taxonomy.rds")

# Create a combined region-disease label
chatila_micro$Region_disease <- paste(chatila_micro$Region, chatila_micro$Disease, sep = "_")
chatila_micro <- AddMetaData(chatila_micro, metadata = chatila_micro$Region_disease, col.name = "Region_disease")

# Subset to specific brain regions
Idents(chatila_micro) <- "Region_disease"
chatila_obj <- subset(chatila_micro, idents = c("SN_NPC", "VTA_NPC", "SI_NPC", "HypoTs_NPC"))

# Update and clean metadata
chatila_obj$cluster_label <- chatila_obj$final_names
chatila_obj$donor_id <- chatila_obj$Donor
chatila_obj$Status <- chatila_obj$Region
chatila_obj$dataset <- "chatila_etal"
chatila_obj$Donor <- NULL

# Export metadata
meta_chatila <- chatila_obj@meta.data[, c("sample_ID", "batch", "Region", "Region_disease", "Sex", "donor_id", "cluster_label", "Age")]
write.csv(meta_chatila, "meta_chatila_microglia.csv", row.names = FALSE)

# Remove layers if present
chatila_obj@assays$RNA@layers$data <- NULL
chatila_obj@assays$RNA@layers$scale.data <- NULL

cat("chatila_etal object processed.\n")
```

# Load and preprocess siletti_etal microglia data
```{r}
cat("Loading siletti_etal microglia data...\n")

siletti_counts <- readRDS("Counts_siletti_etal_micro.RDS")
siletti_meta <- readRDS("meta_siletti_etal_micro.RDS")

siletti_obj <- CreateSeuratObject(counts = siletti_counts, meta.data = siletti_meta)
siletti_obj$Region <- siletti_obj$tissue
siletti_obj$tissue <- NULL
siletti_obj$dataset <- "siletti_etal"

# Remove donor H18.30.001 as this donor only has nuclei from cerebral cortex
siletti_obj <- subset(siletti_obj, subset = donor_id != "H18.30.001")

cat("siletti_etal object processed.\n")
```

# Integrate datasets
```{r}
cat("Beginning dataset integration...\n")

# Combine into list
obj.list <- list(chatila_obj, siletti_obj)

# Normalize and find variable features
obj.list <- lapply(obj.list, function(x) {
  x <- NormalizeData(x)
  x <- FindVariableFeatures(x, selection.method = "vst", nfeatures = 2000)
  return(x)
})

# Select features and anchors
features <- SelectIntegrationFeatures(object.list = obj.list)
anchors <- FindIntegrationAnchors(object.list = obj.list, anchor.features = features)

# Integrate datasets
integrated_microglia <- IntegrateData(anchorset = anchors, k.weight = 25)
cat("Integration complete.\n")
```

# Dimensionality reduction and clustering
```{r}
integrated_microglia <- ScaleData(integrated_microglia, vars.to.regress = "nCount_RNA")
DefaultAssay(integrated_microglia) <- "integrated"

integrated_microglia <- RunPCA(integrated_microglia, npcs = 30, verbose = FALSE)
integrated_microglia <- FindNeighbors(integrated_microglia, dims = 1:22)
integrated_microglia <- FindClusters(integrated_microglia, resolution = c(0.2, 0.3, 0.4, 0.5))
integrated_microglia <- RunUMAP(integrated_microglia, reduction = "pca", dims = 1:22)
```

```{r}
Idents(integrated_microglia) <- "integrated_snn_res.0.2"
DotPlot(integrated_microglia, features = c("CD83", "BCAS2", "DNAJB1", "SLC7A5", "CD69", "HIF1A"))
```

The integrated cluster 6 at resolution 0.2 carries the CD83_BCAS2 signature. 

# Save final object
```{r}
saveRDS(integrated_microglia, "integrated_microglia_chatila_siletti_etal.rds")
cat("Integrated microglia object saved.\n")
```

# Caluclate siletti proportions
## Get props function
```{r}
get_props <- function(data, id, cluster, supercluster = NULL,
                      by = NULL, add_supercluster_prop = F,
                      add_other_cols = T) {
  
  if (!is.data.frame(data)) {
    stop("Input 'data' must be a data frame.")
  }
  
  if (add_supercluster_prop & is.null(supercluster)) {
    stop("No supercluster column provided when add_supercluster_prop == TRUE.")
  }
  
  # Respect existing groups when called in a dplyr pipeline
  if (dplyr::is_grouped_df(data)) {
    if (is.null(by)) {
      by <- dplyr::group_vars(data)
    }
    data <- dplyr::ungroup(data)
  }
  
  groups <- c(id, cluster)
  if (!is.null(supercluster)) {
    groups <- c(groups, supercluster)
  }
  if (!is.null(by)) {
    groups <- c(groups, by)
  }
  
  missing_cols <- setdiff(groups, colnames(data))
  if (length(missing_cols) > 0) {
    stop("The following columns are missing in the data: ",
         paste(missing_cols, collapse = ", "))
  }
  
  # Count cells and compute proportions
  margins <- setdiff(groups, cluster)
  n_cluster <- table(data[groups])
  prop <- as.data.frame(prop.table(n_cluster, margin = margins))
  n_cluster <- as.data.frame(n_cluster)
  colnames(prop)[colnames(prop) == "Freq"] <- "prop"
  colnames(n_cluster)[colnames(n_cluster) == "Freq"] <- paste0("num_", cluster)
  props <- merge(n_cluster, prop, by = groups)

  # Exclude non-existing combinations
  props <- props[!is.nan(props$prop) & !is.nan(props$prop), ]
  
  if (!is.null(supercluster) & add_supercluster_prop) {
    # Count cells per id across superclusters and compute proportions
    if(supercluster %in% groups) {
      groups_supercluster <- setdiff(groups, cluster)
    } else {
      # For consistency, ensure id and supercluster are the first two variables
      groups_supercluster <- c(id, supercluster, setdiff(groups, c(id, cluster)))
    }
    margins_supercluster <- setdiff(groups_supercluster, supercluster)
    n_supercluster <- table(data[groups_supercluster])
    prop_supercluster <- as.data.frame(prop.table(
      n_supercluster, margin = margins_supercluster))
    n_supercluster <- as.data.frame(n_supercluster)
    colnames(prop_supercluster)[colnames(prop_supercluster) == "Freq"] <- "prop"
    props_supercluster <- merge(n_supercluster, prop_supercluster,
                                by = groups_supercluster)
    
    # Exclude non-existing combinations
    props_supercluster <- props_supercluster[
      !is.nan(props_supercluster$prop) & !is.nan(props_supercluster$prop), ]
    
    # Add supercluster proportions as tho they are cluster props
    colnames(props_supercluster)[colnames(props_supercluster) == "Freq"] <-
      paste0("num_", cluster)
    colnames(props_supercluster)[
      colnames(props_supercluster) == supercluster] <- cluster
    props_supercluster[supercluster] <- supercluster
    props <- rbind(props, props_supercluster, make.row.names = F)
  }
  
  # Convert columns to original class
  props[[id]] <- as(props[[id]], class(data[[id]]))
  props[[cluster]] <- as(props[[cluster]], class(data[[cluster]]))
  if (!is.null(supercluster)) {
    props[[supercluster]] <- as(props[[supercluster]],
                                class(data[[supercluster]]))
  }
  
  # Add other columns if required, excluding grouping columns
  if (add_other_cols) {
    other_columns <- setdiff(names(data), groups)
    if (length(other_columns) > 0) {
      df <- summarize_by_group(data, id, other_columns)
      props <- merge(props, df, by = id)
    }
  }
  
  # *Validation Checks*
  
  ## Ensure no new non-zero id-cluster combinations were created
  original_combs <- dplyr::distinct(data[groups])
  result_combs <- props[props[paste0("num_", cluster)] > 0, groups]
  if (!is.null(supercluster) & add_supercluster_prop) {
    result_combs <- result_combs[result_combs[supercluster] != supercluster,
                                 groups]
  }
  new_combs <- dplyr::anti_join(result_combs, original_combs,
                                by = groups)

  if (nrow(new_combs) > 0) {
    stop("Unexpected id x cluster x by x (supercluster) combinations.")
  }

  ## Ensure proportions sum to ~1 (allow for minor floating-point errors)
  props_sum <- props %>%
    dplyr::group_by(!!!rlang::syms(margins)) %>%
    dplyr::summarize(n = round(sum(prop), 5), .groups = "drop") %>%
    dplyr::filter(n != 0) %>%
    dplyr::pull(n) %>%
    unique()

  if (any(abs(props_sum - 1) > 0.001)) {
    stop("Proportions do not sum to 1.")
  }
  
  ## Ensure supercluster proportions sum to 1 per id
  if (!is.null(supercluster) && add_supercluster_prop) {
    supercluster_props_sum <- props %>%
      dplyr::filter(!!rlang::sym(supercluster) == !!supercluster) %>%
      dplyr::group_by(!!!rlang::syms(groups_supercluster)) %>%
      dplyr::summarize(n = round(sum(prop), 5), .groups = "drop") %>%
      dplyr::pull(n) %>%
      unique()
    
    if (any(abs(supercluster_props_sum - 1) > 0.001)) {
      stop("Supercluster proportions do not sum to 1.")
    }
  }
  
  ## Ensure id, cluster and supercluster levels are preserved
  for (col in groups) {
    if (!is.null(col) && any(!unique(data[[col]]) %in% unique(props[[col]]))) {
      stop("Column '", col, "' lost some unique levels in the output.")
    }
  }
  
  ## Ensure all by groups are preserved
  if (!is.null(by)) {
    input_groups <- dplyr::distinct(data, !!!rlang::syms(by))
    output_groups <- dplyr::distinct(props, !!!rlang::syms(by))

    if (nrow(dplyr::anti_join(input_groups, output_groups, by = by)) > 0) {
      stop("Some groups from 'by' are missing in the output.")
    }
  }
  
  return(props)
}
```

## Caluclate proportions
```{r}
region_levels <- c(
  "SN", "cerebellum", "hypothalamus", "midbrain", "myelencephalon", "pons", "thalamic complex",
  "cerebral cortex", "cerebral nuclei", "hippocampal formation"
)

props_siletti <- meta %>%
  group_by(donor_id) %>%
  filter(n() > 100) %>%
  ungroup() %>%
  # keep only SN from Columbia dataset and remove spinal cord from Siletti
  filter(!Region %in% c("HypoTs", "VTA", "SI", "spinal cord")) %>%
  select(c("donor_id", "integrated_snn_res.0.2", "Region", "dataset", "Sex",
           "sex", "Age", "age")) %>%
  group_by(Region, dataset) %>%
  get_props("donor_id", "integrated_snn_res.0.2") %>%
  ungroup() %>%
    mutate(
    integrated_snn_res.0.2 = as.character(integrated_snn_res.0.2),
    Sex_combined = coalesce(sex, Sex),
    Sex_combined = case_when(
      sex == "female" ~ "F",
      sex == "male" ~ "M",
      TRUE ~ Sex),
    Age_combined = coalesce(age, Age),
    Region = as.factor(Region),
    Region = factor(Region, levels = region_levels)
    )
  )

glimpse(props_siletti)

# Define cell types
cell_types <- unique(props_siletti$integrated_snn_res.0.2)
cell_types <- structure(cell_types, names = cell_types)
```

# Proportion analysis for CD83_BCAS2 nuclei in siletti data
```{r}
# filter for microglia carrying the CD83_BCAS2 signature
props_siletti6 <- props_siletti %>%
  filter(integrated_snn_res.0.2 == "6") %>%
  droplevels()

  # define  model for assessing CD83_BCAS2 proportions in each region, with SN from our dataset as the reference
  region_diff_siletti <- nlme::lme(
    sqrt(prop) ~ Region + Sex_combined,
    random = ~ 1 | donor_id,
    data = props_siletti6
  )
  
  # create output table with the effect of region
  output <- broom.mixed::tidy(region_diff_siletti, effects = "fixed",
                              conf.int = T)
  write.csv(output, "nlme_siletti.csv")
```


