---
title: "Smajic Validation"
output: html_notebook
author: "Zena Chatila (zkc2001@cumc.columbia.edu)"
output:
  html_document: 
    toc: true
    number_sections: true
editor_options:
  chunk_output_type: console
---

Purpose: The purpose of this notebook is to:
1) Integrate the microlgial nuclei from our dataset with all microlial nuclei in the Smajic et al. (2022) dataset
2) Identify microglia in the Smajic et al. dataset carrying the CD83_BCAS2 signature
3) Perform proportion analysis to identify whether CD83_BCAS2 proportions differ in Parkinson's versus control donors from the Smajic et al. dataset.

# Set Up
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packges
```{r}
library(DT)
library(knitr)
library(Matrix)
library(patchwork)
library(dplyr)
library(Seurat)
library(ggplot2)
library(AnnotationDbi)
library(org.Hs.eg.db)
# source get_props function for calculating proportions
source("./get_props.R")
# source robust_glm function for proportion analysis
source("./robust_glm.R")
```

# Extract microglia from Smajic dataset 
```{r}
smajic <- readRDS("<path_to_datasetA>/input/toplevel_smajic_object.rds")
Idents(smajic) <- "cell_ontology"
DefaultAssay(smajic) <- "RNA"
smajic_microglia <- subset(smajic, idents = "Microglia")

smajic_counts <- smajic_microglia@assays$RNA@counts
smajic_meta <- smajic_microglia@meta.data

saveRDS(smajic_counts, file = "<output_path>/smajic_counts.RDS")
saveRDS(smajic_meta, file = "<output_path>/smajic_meta.RDS")
```

# Convert Ensembl IDs to gene symbols
```{r}
ensembl_ids <- rownames(smajic_counts)
gene_symbols <- mapIds(org.Hs.eg.db,
                       keys = ensembl_ids,
                       column = "SYMBOL",
                       keytype = "ENSEMBL",
                       multiVals = "first")
gene_symbols_unique <- make.unique(ifelse(is.na(gene_symbols), ensembl_ids, gene_symbols))
rownames(smajic_counts) <- gene_symbols_unique

saveRDS(smajic_counts, file = "<output_path>/smajic_counts_converted_geneids.RDS")
```

# Prepare Smajic Seurat object
```{r}
smajic_meta <- readRDS("<output_path>/smajic_meta.RDS")
smajic_counts <- readRDS("<output_path>/smajic_counts_converted_geneids.RDS")

smajic_obj <- CreateSeuratObject(smajic_counts, meta.data = smajic_meta)
smajic_obj$Region <- "smajic_midbrain"
smajic_obj$dataset <- "smajic_etal"
smajic_obj$Region_disease <- paste("midbrain", smajic_obj$condition, sep = "_")
smajic_obj$donor_id <- smajic_obj$patient
smajic_obj$patient <- NULL

DefaultAssay(smajic_obj) <- "RNA"
```

# Prepare Chatila Seurat object
```{r}
chatila_obj <- readRDS("<path_to_datasetB>/input/chatila_microglia_final.rds")
new_column <- paste(chatila_obj$Region, chatila_obj$Disease, sep = "_")
chatila_obj <- AddMetaData(object = chatila_obj, metadata = new_column, col.name = "Region_disease")
chatila_obj$zena_cluster_name <- chatila_obj$final_names

chatila_obj@assays$RNA@layers$data <- NULL
chatila_obj@assays$RNA@layers$scale.data <- NULL
chatila_obj$donor_id <- chatila_obj$Donor
chatila_obj$Donor <- NULL
chatila_obj$dataset <- "chatila_etal"
```

# Merge and process combined object
```{r}
combined_obj <- merge(smajic_obj, chatila_obj)
combined_obj <- NormalizeData(combined_obj)
combined_obj <- FindVariableFeatures(combined_obj)
combined_obj <- ScaleData(combined_obj)
combined_obj <- RunPCA(combined_obj)
```

# Integration using CCA
```{r}
combined_obj <- IntegrateLayers(object = combined_obj,
                                method = CCAIntegration,
                                orig.reduction = "pca",
                                new.reduction = "integrated.cca",
                                verbose = FALSE)

combined_obj <- FindNeighbors(object = combined_obj, dims = 1:30, reduction = "integrated.cca")
combined_obj <- FindClusters(object = combined_obj, resolution = c(0.2, 0.3, 0.4, 0.5))
combined_obj <- RunUMAP(object = combined_obj, reduction = "integrated.cca", dims = 1:30)
```

# Save processed Seurat object
```{r}
saveRDS(combined_obj, file = "<output_path>/umap_bydonor_smajic_chatila_lognorm_cca_micro.rds")
```

# Load metadata
```{r}
data_file <- "smajic.tsv.gz"
if (file.exists(data_file)) {
  meta <- vroom::vroom(data_file)
} else {
  smajic <- readRDS("umap_bydataset_smajic_zena_cca_micro.rds")
  vroom::vroom_write(smajic@meta.data, data_file)
}
```

# Compute props for Smajic data
```{r}
props_smajic <- meta %>%
  group_by(donor_id) %>%
  filter(dataset == "smajic_etal") %>%
  filter(n() > 100) %>%
  ungroup() %>%
  mutate(across(starts_with("RNA_snn_res.0."), as.character)) %>%
  select(c("donor_id", "RNA_snn_res.0.2", "condition",
                   "sex", "age_at_death")) %>%
  get_props("donor_id", "RNA_snn_res.0.2") %>%
  mutate(condition = as.factor(condition),
         condition = factor(condition, levels = c("Control", "IPD")),
         sex = as.factor(sex),
         sex = factor(sex, levels = c("M", "F")))
glimpse(props_smajic)

# Define cell types
cell_types <- unique(props_smajic$RNA_snn_res.0.2)
cell_types <- structure(cell_types, names = cell_types)
```

# Prop analysis for Smajic data
• Integrated cluster 10 (res 0.2) carries the CD83_BCAS2 Signature
```{r}
# SN comparison: robust GLM of SN
comp <- purrr::map_dfr(cell_types, function(cell_type) {
  output <- robust_glm(prop ~ condition + age_at_death + sex,
                       subset(props_smajic, RNA_snn_res.0.2 == cell_type),
                       family = "quasibinomial")
  tidy_terms(output, exponentiate = F)
}, .id = "cell_type") %>%
  group_by(term) %>%
  mutate(fdr = p.adjust(p.value_hc, "fdr")) %>%
  ungroup()

# how many clusters are enriched/depeleted in disease?
comp %>%
  mutate(cell_type = as.factor(cell_type), term = as.factor(term)) %>%
  datatable()
```
