---
title: "R Notebook"
output: html_notebook
---

```{r}
library(Seurat)
library(dplyr)
library(ggplot2)
library(tidyverse)
```

```{r}
clustered_siletti <- readRDS("clustered_bydataset_Siletti_zena_lognorm_microglia_integrated.rds")
```
```{r}
head(clustered_siletti@meta.data)
```
```{r}
clustered_siletti$new_zena_annotation <- clustered_siletti$zena_cluster_name
clustered_siletti$new_zena_annotation <- gsub("Zena_0", "M1", clustered_siletti$new_zena_annotation)
clustered_siletti$new_zena_annotation <- gsub("Zena_1", "M2", clustered_siletti$new_zena_annotation)
clustered_siletti$new_zena_annotation <- gsub("Zena_2", "M3", clustered_siletti$new_zena_annotation)
clustered_siletti$new_zena_annotation <- gsub("Zena_3", "M4", clustered_siletti$new_zena_annotation)
clustered_siletti$new_zena_annotation <- gsub("Zena_4", "M5", clustered_siletti$new_zena_annotation)
clustered_siletti$new_zena_annotation <- gsub("Zena_5", "M6", clustered_siletti$new_zena_annotation)
clustered_siletti$new_zena_annotation <- gsub("Zena_6", "M7", clustered_siletti$new_zena_annotation)
clustered_siletti$new_zena_annotation <- gsub("Zena_7", "PVM", clustered_siletti$new_zena_annotation)
clustered_siletti$new_zena_annotation <- gsub("Zena_8", "M8", clustered_siletti$new_zena_annotation)
clustered_siletti$new_zena_annotation <- gsub("Zena_9", "M9", clustered_siletti$new_zena_annotation)
clustered_siletti$new_zena_annotation <- gsub("Zena_10", "M10", clustered_siletti$new_zena_annotation)
clustered_siletti$new_zena_annotation <- gsub("Zena_11", "M11", clustered_siletti$new_zena_annotation)
clustered_siletti$new_zena_annotation <- gsub("Zena_12", "M12", clustered_siletti$new_zena_annotation)
clustered_siletti$new_zena_annotation <- gsub("Zena_13", "M13", clustered_siletti$new_zena_annotation)
clustered_siletti$new_zena_annotation <- gsub("Zena_14", "Mono", clustered_siletti$new_zena_annotation)

```


```{r}
clustered_siletti$new_region <- clustered_siletti$region
clustered_siletti$new_region <- gsub("SNCTRL", "SN_NPC", clustered_siletti$new_region)

```

```{r}

```


```{r}
p=DimPlot(clustered_siletti, group.by = "new_region")+NoLegend()
ggsave(p,file="Siletti_byregion_nolegned.tiff")

#DimPlot(clustered_siletti, group.by = "region",split.by = "region")+NoLegend()


Idents(clustered_siletti) <- "new_zena_annotation"
p=DimPlot(clustered_siletti, cells.highlight = WhichCells(clustered_siletti, idents = "M8"), label = T) +
  scale_color_manual(labels = c("Unseleted", "M8"), values = c("grey", "Dark red")) +
  labs(color = "legend title")+NoLegend()
ggsave(p,file="M8_highlighted_siletti_nolegend.tiff")

p=DimPlot(clustered_siletti, group.by = "integrated_snn_res.0.4", label = T)+NoLegend()
ggsave(p,file="Siletti_integrated_clusters_nolegend.tiff")
```
```{r}
Idents(clustered_siletti) <- "region"
DimPlot(clustered_siletti, label=F, cells.highlight=WhichCells(clustered_siletti, idents = "midbrain"), cols.highlight = "darkred", cols= "grey")
DimPlot(clustered_siletti, label=F, cells.highlight=WhichCells(clustered_siletti, idents = "SNCTRL"), cols.highlight = "darkred", cols= "grey")
Idents(clustered_siletti) <- "dissection"

DimPlot(clustered_siletti, label=F, cells.highlight=WhichCells(clustered_siletti, idents = "Midbrain (M) - Substantia Nigra - SN"), cols.highlight = "darkred", cols= "grey")

```


```{r}

#Idents(clustered_siletti) <- "zena_cluster_name"
#DimPlot(clustered_siletti, label=T, group.by="zena_cluster_name", cells.highlight=WhichCells(clustered_siletti, idents = "Zena_8"), cols.highlight = "darkred", cols= "grey")
#Idents(clustered_siletti) <- "dis"
#DimPlot(clustered_siletti, label=T, group.by="region", cells.highlight=WhichCells(clustered_siletti, idents = "midbrain"), cols.highlight = "darkred", cols= "grey")
```

```{r}
Idents(clustered_siletti) <- "integrated_snn_res.0.4"

Idents(clustered_siletti) <- factor(clustered_siletti@active.ident, 
                            levels=c("0","1", "2", "3", "4","5","6","7","8","9","10","11","12","13","14"))
```


```{r}

DefaultAssay(clustered_siletti) <- "RNA"
#Idents(clustered_siletti) <- "integrated_snn_res.0.4"
p=DotPlot(clustered_siletti, features = c("CD83","HSPH1", "HSP90AA1", "HIF1A", "HSPB1", "GPR183","BCAS2", "NAMPT", "TRA2B"), cols = c("lightgrey", "red"))+RotatedAxis()
ggsave(p, file="integratedSiletticlusters_dotplot.tiff", height = 4, width = 6)
```


```{r}
table(clustered_siletti$integrated_snn_res.0.4, clustered_siletti$zena_cluster_name)
table(clustered_siletti$integrated_snn_res.0.4, clustered_siletti$sex)

```

```{r}
table(clustered_siletti$donor_id)
#table(clustered_siletti$donor_id,clustered_siletti$region)
table(clustered_siletti$integrated_snn_res.0.4,clustered_siletti$region)
table(clustered_siletti$donor_id,clustered_siletti$region)

```

```{r}
df2_combined <- df2_combined%>%
  group_by(donor,region)%>%
  mutate(fraction = frequency/sum(frequency))

ggplot(df2_combined, aes(x=cluster, y=fraction, fill=region)) + 
  geom_boxplot()+RotatedAxis()

ggplot(df2_combined, aes(x=cluster, y=log(fraction), fill=region)) + 
  geom_boxplot()+RotatedAxis()
```


```{r}
Idents(clustered_siletti) <- "dataset"
columbia <- subset(clustered_siletti, idents = "Columbia")
siletti <- subset(clustered_siletti, idents = "Siletti_etal")

df_columbia <- as.data.frame(table(columbia$donor_id,columbia$region, columbia$integrated_snn_res.0.4))#, label_transfer_Kamath$Status))
colnames(df_columbia)<-c("donor","region", "cluster", "frequency")

df_siletti <- as.data.frame(table(siletti$donor_id,siletti$region, siletti$integrated_snn_res.0.4))#, label_transfer_Kamath$Status))
colnames(df_siletti)<-c("donor","region", "cluster", "frequency")

df2_combined <- rbind(df_columbia, df_siletti)

df2_combined <- df2_combined%>%
  group_by(donor,region)%>% #frequency=no. of nuclei per donor/region/cluster 
  mutate(fraction = frequency/sum(frequency)) ##  freq/total nuclei for the donor

ggplot(df2_combined, aes(x=cluster, y=fraction, fill=region)) + 
  geom_boxplot()+RotatedAxis()

ggplot(df2_combined, aes(x=cluster, y=log(fraction), fill=region)) + 
  geom_boxplot()+RotatedAxis()

```
```{r}
df_combined_c7 <- df2_combined[df2_combined$cluster=="7",]

ggplot(df_combined_c7, aes(x=cluster, y=fraction, fill=region)) + 
  geom_boxplot()+RotatedAxis()

ggplot(df_combined_c7, aes(x=cluster, y=log(fraction), fill=region)) + 
  geom_boxplot()+RotatedAxis()

```
```{r}
write.csv(df_combined_c7, file = "Siletti_boxplot_table.csv")
```


```{r}
p=ggplot(df_combined_c7, aes(x=cluster, y=fraction, fill=region)) + 
  geom_boxplot()+RotatedAxis()+theme_bw()+
  
  scale_fill_brewer(name = "region",palette="Set1")+
  scale_y_continuous(trans="log10")+
  ylab("Proportion")+
  xlab("")#+
  #theme(legend.position = c(0.9255, 0.8),
    #    legend.background = element_rect(
     #     size=0.5, linetype="solid", 
    #      color ="black"))

ggsave(p, file="box.png")
```


####prepare for MASC


```{r}
table(clustered_siletti$age)
#table(clustered_siletti$combined_age, clustered_siletti$dataset)

```

##MASC function below
```{r}
#run masc
#library(tidyverse)
require(lme4)
library(dplyr)
#masc function
#############
MASC <- function(dataset, cluster, contrast, random_effects,
                 fixed_effects = NULL, verbose = FALSE, save_models = FALSE,
                 save_model_dir = NULL){
  if(is.factor(dataset[[contrast]]) == FALSE) {
    stop("Specified contrast term is not coded as a factor in dataset")
  }
  cluster <- as.character(cluster)
  designmat <- model.matrix(~cluster + 0, data.frame(cluster = cluster))
  dataset <- cbind(designmat, dataset)
  cluster <- as.character(cluster)
  designmat <- model.matrix(~cluster + 0, data.frame(cluster = cluster))
  dataset <- cbind(designmat, dataset)
  res <- vector(mode = "list", length = length(unique(cluster)))
  names(res) <- attributes(designmat)$dimnames[[2]]
  if(!is.null(fixed_effects) && !is.null(random_effects)){
    model_rhs <- paste0(c(paste0(fixed_effects, collapse = " + "),
                          paste0("(1|", random_effects, ")", collapse = " + ")),
                        collapse = " + ")
    if(verbose == TRUE) {
      message(paste("Using null model:", "cluster ~", model_rhs))
    }
  } else if(!is.null(fixed_effects) && is.null(random_effects)){
    model_rhs <- paste0(fixed_effects, collapse = " + ")
    if (verbose == TRUE) {
      message(paste("Using null model:", "cluster ~", model_rhs))
      stop("No random effects specified")
    }
  } else if(is.null(fixed_effects) && !is.null(random_effects)){
    model_rhs <- paste0("(1|", random_effects, ")", collapse = " + ")
    if (verbose == TRUE) {
      message(paste("Using null model:", "cluster ~", model_rhs))
    }
  } else{
    model_rhs <- "1"
    if(verbose == TRUE) {
      message(paste("Using null model:", "cluster ~", model_rhs))
      stop("No random or fixed effects specified")
    }
  }
  cluster_models <- vector(mode = "list", length = length(attributes(designmat)$dimnames[[2]]))
  names(cluster_models) <- attributes(designmat)$dimnames[[2]]
  print("Beginning Modelling")
  for(i in seq_along(attributes(designmat)$dimnames[[2]])) {
    test_cluster <- attributes(designmat)$dimnames[[2]][i]
    if(verbose == TRUE) {
      message(paste("Creating logistic mixed models for",
                    test_cluster))
    }
    dataset[,i] <- factor(dataset[,i])
    null_fm <- as.formula(paste0(c(paste0(test_cluster, " ~ 1 + "),
                                   model_rhs), collapse = ""))
    full_fm <- as.formula(paste0(c(paste0(test_cluster, " ~ ",
                                          contrast, " + "), model_rhs), collapse = ""))
    null_model <- lme4::glmer(formula = null_fm, data = dataset,
                              family = binomial, nAGQ = 1, verbose = 0, control = glmerControl(optimizer = "bobyqa"))
    full_model <- lme4::glmer(formula = full_fm, data = dataset,
                              family = binomial, nAGQ = 1, verbose = 0, control = glmerControl(optimizer = "bobyqa"))
    model_lrt <- anova(null_model, full_model)
    contrast_lvl2 <- paste0(contrast, levels(dataset[[contrast]])[2])
    contrast_ci <- confint.merMod(full_model, method = "Wald",
                                  parm = contrast_lvl2)
    cluster_models[[i]]$null_model <- null_model
    cluster_models[[i]]$full_model <- full_model
    cluster_models[[i]]$model_lrt <- model_lrt
    cluster_models[[i]]$confint <- contrast_ci
  }
  output <- data.frame(cluster = attributes(designmat)$dimnames[[2]],
                       size = colSums(designmat))
  output$model.pvalue <- sapply(cluster_models, function(x) x$model_lrt[["Pr(>Chisq)"]][2])
  output[[paste(contrast_lvl2, "OR", sep = ".")]] <- sapply(cluster_models,
                                                            function(x) exp(fixef(x$full)[[contrast_lvl2]]))
  output[[paste(contrast_lvl2, "OR", "95pct.ci.lower", sep = ".")]] <- sapply(cluster_models,
                                                                              function(x) exp(x$confint[contrast_lvl2, "2.5 %"]))
  output[[paste(contrast_lvl2, "OR", "95pct.ci.upper", sep = ".")]] <- sapply(cluster_models,
                                                                              function(x) exp(x$confint[contrast_lvl2, "97.5 %"]))
  if(save_models == TRUE){
    saveModelObj(cluster_models, save_dir = save_model_dir)
    return(output)
  }
  else{
    return(output)
  }
}
####################
```

```{r}
metadata <- clustered_siletti@meta.data
```



```{r}
table(metadata$region)
```


```{r}
# Define a vector of regions to include (not including )
metadata$region <- gsub(" ", "_",metadata$region)

regions <- c("cerebellum","cerebral_cortex", "cerebral_nuclei", "midbrain", "hippocampal_formation", "hypothalamus","myelencephalon", "pons", "thalamic_complex")

# Loop over regions
for (region in regions) {
  # Define the keepindices for this region and SNCTRL
  keepindices <- which(metadata$region %in% c(region, "SNCTRL"))
  
  # Subset the metadata for this region and SNCTRL
  meta <- metadata[keepindices,]
  meta$region <- factor(meta$region)
  #meta$sex <- factor(meta$sex)
  #meta$age <- factor(meta$age)
  meta$donor_id <- factor(meta$donor_id)
  
  # Run the MASC analysis for this region and SNCTRL
  MASC_res_region_SNCTRL <- MASC(data=meta, cluster=meta$integrated_snn_res.0.4, 
                                 contrast = "region", random_effects = "donor_id",
                                 verbose=T) %>%
    dplyr::mutate(fdr = p.adjust(model.pvalue, method = "fdr")) %>%
    dplyr::arrange(model.pvalue)
  
  # Assign the results to a variable with a name based on the region
  assign(paste0("noSex_MASC_res_", region, "_vs_SNCTRL"), MASC_res_region_SNCTRL)
}
```
```{r}
# Define a vector of regions to include (not including )
metadata$region <- gsub(" ", "_",metadata$region)

regions <- c("cerebellum","cerebral_cortex", "cerebral_nuclei", "midbrain", "hippocampal_formation", "hypothalamus", "thalamic_complex")

# Loop over regions
for (region in regions) {
  # Define the keepindices for this region and SNCTRL
  keepindices <- which(metadata$region %in% c(region, "SNCTRL"))
  
  # Subset the metadata for this region and SNCTRL
  meta <- metadata[keepindices,]
  meta$region <- factor(meta$region)
  #meta$sex <- factor(meta$sex)
  #meta$age <- factor(meta$age)
  meta$donor_id <- factor(meta$donor_id)
  
  # Run the MASC analysis for this region and SNCTRL
  MASC_res_region_SNCTRL <- MASC(data=meta, cluster=meta$integrated_snn_res.0.4, 
                                 contrast = "region", random_effects = "donor_id",
                                 verbose=T) %>%
    dplyr::mutate(fdr = p.adjust(model.pvalue, method = "fdr")) %>%
    dplyr::arrange(model.pvalue)
  
  # Assign the results to a variable with a name based on the region
  assign(paste0("noSex_MASC_res_", region, "_vs_SNCTRL"), MASC_res_region_SNCTRL)
}
```

###combine the output of MASC files and extract out cluster 7 from all these outputs and make a new table. in teh below code I fisrt merge the results wich had same column. THalamic complex had diff column name os that was merged later
```{r}
noSex_MASC_res_midbrain_vs_SNCTRL
# Define a list of data frame names to process
df_list <- list("noSex_MASC_res_hippocampal_formation_vs_SNCTRL", "noSex_MASC_res_cerebellum_vs_SNCTRL", "noSex_MASC_res_cerebral_cortex_vs_SNCTRL", "noSex_MASC_res_cerebral_nuclei_vs_SNCTRL", "noSex_MASC_res_hypothalamus_vs_SNCTRL", "noSex_MASC_res_midbrain_vs_SNCTRL","noSex_MASC_res_myelencephalon_vs_SNCTRL"
                , "noSex_MASC_res_pons_vs_SNCTRL")

# Create an empty list to store the results
results_list <- list()

# Loop through each data frame and extract the "cluster7" row
for (df_name in df_list) {
  # Load the data frame
  df <- get(df_name)
  
  # Extract the "cluster7" row and store it in the list
  results_list[[df_name]] <- df %>%
    filter(cluster == "cluster7") %>%
    select(-cluster)
}

# Combine the list of results into a single data frame
results_df <- do.call(rbind, results_list)
```


```{r}
results_df$region <- rownames(results_df)
results_df$upper <- results_df$regionSNCTRL.OR.95pct.ci.upper
results_df$lower <- results_df$regionSNCTRL.OR.95pct.ci.lower
results_df$OddsRatio <- results_df$regionSNCTRL.OR

results_df$regionSNCTRL.OR.95pct.ci.upper <- NULL
results_df$regionSNCTRL.OR.95pct.ci.lower <- NULL
results_df$regionSNCTRL.OR <- NULL
```

```{r}
res_thalamic <- noSex_MASC_res_thalamic_complex_vs_SNCTRL %>%
    filter(cluster == "cluster7") %>%
    select(-cluster)
res_thalamic[1,3]= 1/res_thalamic[1,3]
res_thalamic[1,4]= 1/res_thalamic[1,4]
res_thalamic[1,5]= 1/res_thalamic[1,5]

res_thalamic$region <- rownames(res_thalamic)
res_thalamic$lower <- res_thalamic$regionthalamic_complex.OR.95pct.ci.upper
res_thalamic$upper <- res_thalamic$regionthalamic_complex.OR.95pct.ci.lower
res_thalamic$OddsRatio <- res_thalamic$regionthalamic_complex.OR

res_thalamic$regionthalamic_complex.OR.95pct.ci.upper <- NULL
res_thalamic$regionthalamic_complex.OR.95pct.ci.lower <- NULL
res_thalamic$regionthalamic_complex.OR <- NULL
```
```{r}
result_combined_table <- rbind(results_df,res_thalamic)
result_combined_table$region <- c("hippocampal_formation_vs_SNCTRL", "cerebellum_vs_SNCTRL","cerebral_cortex_vs_SNCTRL","cerebral_nuclei_vs_SNCTRL", "hypothalamus_vs_SNCTRL","midbrain_vs_SNCTRL","myelencephalon_vs_SNCTRL", "pons_vs_SNCTRL", "thalamic_compex_vs_SNCTRL")
```

```{r}
write.csv(result_combined_table, file="i-7_pairwise_region_MASC.csv")
```


```{r}

or <- ggplot(data=result_combined_table %>% as_tibble() %>% 
               mutate(region = str_remove(region,"_vs_SNCTRL"),
                      region = str_replace(region,"_","\n")) %>%
               arrange(size) %>%
               mutate(region = factor(region, levels = region),
                      sig = factor(ifelse((log2(upper) >=0) & (log2(lower) <= 0), "longdash","solid"))),
             aes(x=region, y=log2(OddsRatio), ymin=log2(lower), ymax=log2(upper),
                 color=log2(OddsRatio),
                 linetype = sig)) +
  geom_point(aes(size =log2(size))) + 
  geom_errorbar(width=.2,size=.5) +
  labs(title='Other_regions vs SN Control', y='log(OddsRatio)', x = 'Cluster') +# xlim(-10,20) +
  geom_hline(yintercept=0, color='black', linetype='dashed', alpha=.5) +
  scale_color_gradient(low="blue",high="red")+
  scale_linetype_manual(values = c("longdash"="longdash","solid"="solid"))+
  theme_bw()+
  guides(linetype = "none",
         color = "none", size="none")+
  ylab(expression(log[2](OR)))+
  xlab("")+
  theme(axis.text = element_text(size=12))+
  coord_flip()
or


ggsave(or, file="OR.png")


```
```{r}

```




```{r}
#result_combined_table$region <- rownames(result_combined_table)
summary_OR <- sum(result_combined_table$OddsRatio) / nrow(result_combined_table)
result_combined_table |>  
  forestplot( 
           mean=OddsRatio,
            lower = lower, 
           upper = upper, 
           labeltext = region, 
           xlab = "Odds Ratio", 
           col = fpColors(box="black", line="black", summary="black"), 
           boxsize = 0.3,
            xlim = c(-1, 1),
           zero = summary_OR,
           clip = c(-1, 1),
           #txt_gp = fpTxtGp(label = gpar(cex = 0.8)),
           #cex = 0.8,
           #hrzl_lines = list(y = 0, lty = "dashed"),
           #is.summary = c(TRUE, FALSE, FALSE, FALSE),
           #graph.pos = 2,
           #fn.ci_norm = fpDrawNormalCI,
           lwd.ci_norm = 0.5)
         # labeltext= c(cluster,new_subothers.OR),
           #  clip = c(0.1, 2.5))
            # xlog = TRUE)) |>
  
   
```

```{r}
#table(metadata$donor_id,metadata$sex)
table(metadata$region,metadata$age)

```

##Siletti SN
  Midbrain (M) - Inferior colliculus and nearby nuclei - IC                                                                                                                                                 0
  Midbrain (M) - Periaqueductal gray and Dorsal raphe nucleus - PAG-DR                                                                                                                                      0
  Midbrain (M) - Periaqueductal gray and nearby nuclei - PAG                                                                                                                                                0
  Midbrain (M) - Pretectal region - PTR                                                                                                                                                                     0
  Midbrain (M) - Substantia Nigra - SN                                                                                                                                                                      0
  Midbrain (M) - Substantia Nigra and Red Nucleus - SN-RN                                                                                                                                                   0
  Midbrain (M) - Substantia Nigra, Red Nucleus, and nearby nuclei - SN-RN                                                                                                                                   0
  Midbrain (M) - Superior colliculus and nearby nuclei - SC                                                                                                                                                 0
  Midbrain (RN) - Red Nucleus - RN                                                         


```{r}

df_columbia <- as.data.frame(table(columbia$donor_id,columbia$region, columbia$integrated_snn_res.0.4))#, label_transfer_Kamath$Status))
colnames(df_columbia)<-c("donor","region", "cluster", "frequency")


df_siletti_dissection <- as.data.frame(table(siletti$donor_id,siletti$dissection, siletti$integrated_snn_res.0.4))#, label_transfer_Kamath$Status))
colnames(df_siletti_dissection)<-c("donor","region", "cluster", "frequency")



df2_combined <- rbind(df_columbia, df_siletti_dissection)
df2_combined <- df2_combined%>% ## this step needs to be done before subsetting
  group_by(donor,region)%>%
  mutate(fraction = frequency/sum(frequency))

df_combined_c7 <- df2_combined[df2_combined$cluster=="7",]
subset_df7 <- df_combined_c7[grepl("Midbrain|SNCTRL", df_combined_c7$region), ]



```


```{r}


ggplot(subset_df7, aes(x=cluster, y=fraction, fill=region)) + 
  geom_boxplot()+RotatedAxis()

ggplot(subset_df7, aes(x=cluster, y=log(fraction), fill=region)) + 
  geom_boxplot()+RotatedAxis()

```
```{r}
table(clustered_siletti$donor_id)

```

```{r}
Idents(siletti) <- "region"
siletti_midbrain <- subset(siletti, idents = "midbrain")
```

```{r}
table(siletti_midbrain$dissection)
```
















